LaMachine: A meta-distribution for NLP software

### 2019-01-17 docs/lamachine_clarin2018.tex
paper update

LaMachine incorporates software providing different types of interfaces\footnote{Command line interfaces, programming
interfaces, web-user interfaces, webservices.} which, as seen above, typically address different audiences. Whilst we
attempt to accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and
less-technical audiences\footnote{The wider researcher community, particularly the Humanities; also educational
settings.}, there is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
higher-level interfaces on. Depending on the \emph{flavour} of LaMachine chosen, it makes a good virtual research
environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
environment for a developer or a good deployment method for production servers in for example CLARIN centres.  We
demonstrate how LaMachine can create a fully functioning and standalone research environment for text mining  and NLP
for Dutch texts in a use case project at the Healthcare Inspectorate.

@@

\begin{enumerate}
    \item LaMachine facilitates the installation, distribution and configuration of software. It does not fork, modify
        or appropriate the participating software in any way, nor does it provide a hosting place or repository for
        software.
    \item The software included in LaMachine must under a recognised (i.e. OSI-approved) open-source license.
    \item The source
        code must be hosted in a public version controlled repository\footnote{e.g. Github, Gitlab, Bitbucket, provided
        the respository is public}
    \item There must be build process to compile the source, if applicable, and install the program
    \item There must be some kind of release protocol (with semantic versioning) using the proper
technology-specific channels.
    \item All software that is incorporated in LaMachine must bear at least some relevance to the field of Natural Language Processing.
    \item LaMachine lives in an open-source ecosystem and therefore builds on Unix-like platforms; this primarily
        means Linux, as well as BSD and, with some restrictions, macOS. This by definition excludes certain software for different platforms, such as
        mobile platforms (Android/iOS/etc), native Windows/mac desktop applications, or certain interface types in
        general such as classical desktop GUI applications or mobile `apps', all of which fall beyond our scope.
        Cygwin\footnote{A unix environment on Windows} is not tested or
        supported.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows.
\end{enumerate}

@@

\begin{enumerate}
    \item \textbf{Local installation} - This installs the bulk of LaMachine in a separate local virtual environment (a seperate
        directory) that has to be explicitly activated to be used. This also allows for multiple different installations
        of LaMachine on the same host system (e.g. for different users or with different software configurations).
        This local installation still actively relies on various global dependencies that are available through the
        package manager of your distribution.
    \item \textbf{Global installation} - This flavour is used for a host that is fully dedicated to LaMachine. Everything will be
        installed globally so there is only one installation possible, multiple users will all make use of the same
        installation.
    \item \textbf{Docker container} - Installs LaMachine in a container. Containerisation separates the entire run-time
        environment from the host system, only the Linux kernel is shared. It is a lighter option than full
        virtualisation.
    \item \textbf{Singularity container} - Another form of containerisation with more focus on security\footnote{reduced
        privilege escalation}, making it more suitable than Docker for deployment on computing clusters.
    \item \textbf{Virtual Machine} - Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
        deployment of LaMachine on hosts which would otherwise not support it, such as Windows. Virtualisation for
        LaMachine is achieved through Vagrant and VirtualBox\footnote{\url{https://vagrant.org}, \url{https://www.virtualbox.org}}.
    \item \textbf{Remote provisioning} - Installs LaMachine on a remote dedicated server. This option is most suited for
        hosting centres.
\end{enumerate}

@@

The two native options, local and global installation, support a variety of major GNU/Linux distributions: Debian,
Ubuntu, Arch Linux, CentOS, Fedora and, to a more limited degree, we also support macOS (the local installation flavour), powered in part by the homebrew package
manager\footnote{Alternatives such as macports are not supported}. Support for macOS is limited because not all participating
software supports it. Linux Distributions that are derivates of these \emph{may} also work, such as RedHat Enterprise Linux (CentOS), Linux Mint (Ubuntu),
and Manjaro (Arch Linux).

@@

The nomenclature is admittedly a bit confusing, but the notion of version discussed in the section refers to the
versions of the various software packages inside LaMachine. It should not be
confused with the actual \emph{version release number} of Lamachine as a whole, which is the version number assigned to
the collective of installation scripts LaMachine provides, and which marks LaMachine releases.

@@

Installation of LaMachine begins with a single bootstrap command\footnote{See \url{https://proycon.github.io/LaMachine}}
executed on the command line.  It can interactively query the users for their software preferences \emph{(stored as the
host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install, \emph{the installation
manifest}. This set is never static but can be customized by the user. The bootstrap procedure, shown in
Figure~\ref{fig:bootstrap}, detects and installs the
necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work, unless a
pre-published container or VM is selected.  Figure~\ref{fig:arch} provides a schematic view of the LaMachine
architecture.

@@

Once LaMachine is installed, in any of its flavours, you can update it from inside by running
\texttt{lamachine-update}. This updates all of the software managed by LaMachine.

@@

LaMachine aims to harmonise the metadata of all installed software, by converting metadata from upstream repositories,
i.e. the repositories where tool providers deposit their software, to a common yet simple standard called CodeMeta
\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
encouraging software developers to provide their codemeta metadata inside their source code repositories and using that
directly. This in turn enables other tools to do proper service discovery and provenance logging.

@@

\begin{itemize}
 \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University
     Nijmegen\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
    \url{https://languagemachines.github.io/ucto}, \url{https://languagemachines.github.io/frog},
    \url{http://ilk.uvt.nl/wopr}, \url{https://proycon.github.io/folia}, \url{https://proycon.github.io/clam},
\url{https://github.com/proycon/flat}, \url{https://proycon.anaproy.nl/pynlpl},
\url{https://proycon.github.io/colibri-core/},
\url{https://github.com/proycon/gecco},
\url{https://github.com/proycon/valkuil-gecco},
\url{https://github.com/LanguageMachines/PICCL}, \url{https://githu.com/LanguageMachines/ticcltools},
\url{https://github.com/proycon/labirinto},
\url{https://github.com/proycon/oersetter-webservice}
}
 \begin{itemize}
     \item \textbf{Timbl} - A memory-based machine learning
         toolkit, and \textbf{Mbt}, a memory-based tagger based on timbl.
     \item \textbf{Ucto} - A tokeniser
     \item \textbf{Frog} - Frog is an integration of various memory-based natural language processing (NLP) modules
         developed for Dutch. It can do Part-of-Speech tagging, lemmatisation, named entity recogniton, shallow parsing,
         dependency parsing and morphological analysis. Also included in LaMachine;
         \textbf{Toad}, Trainer Of All Data, training tools for Frog.
     \item \textbf{Wopr}- Memory-based Word Predictor.
     \item \textbf{CLAM} - Quickly build RESTful webservices, powers many webservices offered by LaMachine.
     \item \textbf{FoLiA} - Format for Linguistic Annotation, with tools and libraries in/for Python and $C++$.
     \item \textbf{FLAT} - FoLiA Linguistic Annotation Tool: a web-based linguistic annotation tool.
     \item \textbf{PyNLPl} - Python Natural Language Processing Library.
     \item \textbf{Colibri Core} -- Colibri core is an NLP tool as well as a C++ and Python library for working with
         basic linguistic constructions such as n-grams and skipgrams (i.e patterns with one or more gaps, either of
         fixed or dynamic size) in a quick and memory-efficient way. -
     \item \textbf{Gecco} - Generic Environment for Context-Aware Correction of Orthography - and \textbf{Valkuil.net},
         an instantiation thereof for Dutch.
     \item \textbf{PICCL} - PICCL: A set of workflows for corpus building through OCR, post-correction (through \textbf{TICCL}) and Natural Language Processing.
     \item \textbf{Labirinto}- A web-based portal listing all
         available tools in LaMachine, an ideal starting point for LaMachine.
     \item \textbf{Oersetter} - A Frisian-Dutch Machine Translation system in collaboration with the Fryske Akademy.
 \end{itemize}
 \item by the University of Groningen
 \begin{itemize}
     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - a dependency parser and tagger for Dutch
 \end{itemize}
 \item by the Vrije Universiteit Amsterdam
 \begin{itemize}
     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files
 \end{itemize}
 \item by Utrecht University
 \begin{itemize}
     \item  \textbf{T-scan}\footnote{\url{https://github.com/proycon/tscan}} - T-scan is a Dutch text analytics tool for readability prediction (initially developed at TiCC, Tilburg University, and in collaboration with Radboud University, Nijmegen)
 \end{itemize}
 \item by Meertens Instituut
 \begin{itemize}
     \item \textbf{Python Course for the Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} - Interactive
         tutorial and introduction into programming with Python for the humanities
 \end{itemize}
\end{itemize}

In addition to the above listed specific software, LaMachine also incorporates a large number of renowned tools by external international parties, offering
most notably a mature Python environment with renowned scientific modules. The following list is not
exhaustive: \footnote{\url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
    \url{https://spacy.io},  \url{http://www.nextflow.io}, \url{https://stanfordnlp.github.io/CoreNLP/},  \url{https://github.com/tesseract-ocr/tesseract}, \url{https://tensorflow.org},
\url{http://kaldi-asr.org}, \url{http://www.statmt.org/moses}}

\begin{itemize}
    \item \textbf{Python}: Numpy, Scipy, Matplotlib, Scikit-Learn, IPython, Jupyter, ...
    \begin{itemize}
        \item \textbf{Jupyter Lab} The successor of the popular Jupyter Notebooks, offers notebooks, a web-based IDE, terminals. An ideal entry point to get started with LaMachine and all it contains!
        \item \textbf{PyTorch} - Deep-learning library for Python
        \item \textbf{NLTK} - Natural Language Toolkit for Python
        \item \textbf{Spacy} - Industrial-Strength NLP in Python
    \end{itemize}
    \item R
    \item Java
    \begin{itemize}
        \item NextFlow - A system and language for writing parallel and scalable pipelines in a portable manner.
        \item Stanford CoreNLP](https://stanfordnlp.github.io/CoreNLP/) - Various types of linguistic enrichment
    \end{itemize}
    \item \textbf{Tesseract} - Open Source Optical Character Recognition (OCR)
    \item \textbf{Tensorflow} - Open-source machine learning framework
    \item \textbf{Kaldi} - Speech Recognition Framework (ASR)
    \item \textbf{Moses} - Statistical Machine Translation system
\end{itemize}

@@

In any of its flavours, LaMachine offers low-level shell access, i.e. a command-line interface accessed through a
terminal. In flavours that are separated from the host system over the network, this is accomplished over \texttt{ssh}.
At that point, as shown in Figure~\ref{fig:venv}, the user has the liberty to do whatever she wants and can invoke any
of the tools that offer a text interface, including text editors such as vim, emacs or nano, version control systems
such as git and interpreters such as Python or R (or compilers for that matter). The latter in turn allow users to use
one of the many specialised programming libraries included in LaMachine to build their own tools. All this makes
LaMachine ideally suited as a development environment.

@@

LaMachine comes with a webserver\footnote{In situations where web interfaces are not needed or desired, the
user has the ability to opt-out of this}. This webserver serves various web-capable tools that are incorporated in LaMachine. One
of these tools is a portal website that provides an overview of all installed software in
LaMachine, and acts as a point of access for web services and web applications. This portal, as shown in
Figure~\ref{fig:portal}, leverages the metadata registry compiled in each LaMachine installation (see
section~\ref{sec:metadata}).

@@

LaMachine also comes with a
Jupyter Lab\footnote{\url{https://jupyter.org/}} installation which provides a web-based Integrated Development
Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
text, code and data output and have gained great popularity in data science community. This is shown in
Figure~\ref{fig:lab}. This type of virtual laboratory provides a powerful interface that can form an alternative to
regular shell access and may have a larger appeal for those parts of the audience that do not feel completely
comfortable in the terminal.

@@

LaMachine provides various CLAM-based webservices. CLAM is a long-time CLARIN project that allows developers to turn
their command-line NLP tool into a RESTful webservice with relatively little effort \cite{CLAMPAPER}, and which in
addition also provides a generic web-interface for human end-users. The portal website in LaMachine links to these. This
again address an audience that does not require a high degree of technical expertise, but is suitable for a wider
audience of researchers, including those in the Humanities.

@@

Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. But that
would not be appropriate as we characterise LaMachine more as a meta-distribution which simply builds on the foundation
of several existing major Linux distributions. As a meta-distribution that runs in a variety of flavours and on a verity
of platforms, it has a great deal of flexibility that normal distributions don't have. At the same time, the scope of
LaMachine is more narrowly defined than that of a generic Linux distribution, including even specialised Linux
distributions that focus on scientific computing, such as Fedora
Scientific\footnote{\url{https://labs.fedoraproject.org/en/scientific/}}.

@@

At the start of the development cycle for LaMachine v2, we investigated whether Anaconda and its ecosystem would provide
a sufficient foundation for us to build upon. We concluded this was not the case for a number of reasons: Anaconda
introduces more overhead than we desired for our purposes and conflicted with certain technology choices we made. We
wanted our native flavours (local environment and global environment) to be as close to the underlying distribution as possible and to reuse existing
technologies and packages from the distribution (or from homebrew for macOS), such as compilers (e.g. gcc) and
interpreters (e.g. python).  Anaconda, on the other hand, choses to provide its own packages and builds on those. It
does so using its own package manager (conda) and package repositories (conda-forge). This would require repackaging certain software for the
anaconda ecosystem, especially handling of non-Python software was not trivial, as Anaconda was originally conveived as a
Python distribution.

>>>>

LaMachine incorporates software providing different types of interfaces\footnote{Command line interfaces, programming
interfaces, web-user interfaces, webservices.} which, as seen above, typically address different audiences. Whilst we
attempt to accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and
less-technical audiences\footnote{The wider researcher community, particularly the Humanities; also educational
settings.}, there is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
higher-level interfaces on. Depending on the \emph{flavour} (more on this later) of LaMachine chosen, it makes a good virtual research
environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
environment for a developer or a good deployment method for production servers in for example CLARIN centres.  We
demonstrate how LaMachine can create a fully functioning and standalone research environment for text mining  and NLP
for Dutch texts in a use case project at the Healthcare Inspectorate.

@@

\begin{enumerate}
    \item LaMachine facilitates the installation, distribution and configuration of software. It does not fork, modify
        or appropriate the participating software in any way, nor does it provide a hosting place or repository for
        software.
    \item The software included in LaMachine must under a recognised (i.e. OSI-approved) open-source license.
    \item The source
        code must be hosted in a public version controlled repository. \footnote{e.g. Github, Gitlab, Bitbucket, provided
        the respository is public}
    \item There must be a build process to compile the source, if applicable, and install the program or library.
    \item There must be some kind of release protocol (adhering to semantic versioning) that publishes software using the proper
technology-specific channels.
    \item All software that is incorporated in LaMachine must bear at least some relevance to the field of Natural Language Processing.
    \item LaMachine lives in an open-source ecosystem and therefore builds on Unix-like platforms; this primarily
        means Linux, as well as BSD and, with some restrictions, macOS. This by definition excludes certain software for different platforms, such as
        mobile platforms (Android/iOS/etc), native Windows/mac desktop applications, or certain interface types in
        general such as classical desktop GUI applications or mobile `apps', all of which fall beyond our scope.
        Cygwin\footnote{A unix environment on Windows} is not tested or
        supported.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows.
\end{enumerate}

@@

\begin{enumerate}
    \item \textbf{Local installation} - This installs the bulk of LaMachine in a separate local virtual environment (a seperate
        directory) that has to be explicitly activated to be used. This also allows for multiple different installations
        of LaMachine on the same host system (e.g. for different users or with different software configurations).
        This local installation still actively relies on various global dependencies that are available through the
        package manager of your distribution.
    \item \textbf{Global installation} - This flavour is used for a host that is fully dedicated to LaMachine. Everything will be
        installed globally so there is only one installation possible, multiple users will all make use of the same
        installation.
    \item \textbf{Docker container} - Installs LaMachine in a container. Containerisation separates the entire run-time
        environment from the host system, only the Linux kernel is shared. It is a lighter option than full
        virtualisation. \footnote{Note that Docker on other platforms such as Windows and macOS does do use full
        virtualisation.}
    \item \textbf{Singularity container} - Another form of containerisation with more focus on security\footnote{reduced
        privilege escalation}, making it more suitable than Docker for deployment on computing clusters. This flavour is
        still under development.
    \item \textbf{Virtual Machine} - Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
        deployment of LaMachine on hosts which would otherwise not support it, such as Windows. Virtualisation for
        LaMachine is achieved through Vagrant and VirtualBox\footnote{\url{https://vagrant.org}, \url{https://www.virtualbox.org}}.
    \item \textbf{Remote provisioning} - Installs LaMachine on a remote dedicated server. This option is most suited for
        hosting centres.
\end{enumerate}

@@

The two native options, local and global installation, support a variety of major GNU/Linux distributions: Debian,
Ubuntu, Arch Linux, CentOS, Fedora and, to a more limited degree, we also support macOS (the local installation flavour), powered in part by the homebrew package
manager\footnote{\url{https://brew.sh}; alternatives such as macports are not supported}. Support for macOS is limited because not all participating
software supports it. Certain Linux Distributions that are derivatives of these \emph{may} also work, such as RedHat
Enterprise Linux (CentOS) and Linux Mint (Ubuntu).

@@

The nomenclature is admittedly a bit confusing, but the notion of version discussed in this section refers to the
versions of the various software packages inside LaMachine. It should not be
confused with the actual \emph{version release number} of Lamachine as a whole, which is the version number assigned to
the collective of installation scripts LaMachine provides, and which marks LaMachine releases.

@@

Installation of LaMachine begins with a single bootstrap command\footnote{See \url{https://proycon.github.io/LaMachine}}
executed on the command line.  It can interactively query the user for her software preferences \emph{(stored as the
host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install, \emph{the installation
manifest}. This set is never static but can be customised by the user. The bootstrap procedure, a screenshot of which is shown in
Figure~\ref{fig:bootstrap}, detects and installs the
necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work, unless a
pre-published container or VM is selected.  Figure~\ref{fig:arch} provides a schematic view of the LaMachine
architecture.

@@

Once LaMachine is installed, in any of its flavours, it can be updated from inside by running
\texttt{lamachine-update}. This updates all of the software managed by LaMachine.

@@

LaMachine aims to harmonise the metadata of all installed software. This is accomplished by converting metadata from upstream repositories,
i.e. the repositories where tool providers deposit their software, to a common yet simple standard called CodeMeta
\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
encouraging software developers to provide their codemeta metadata inside their source code repositories and using that
directly. This in turn enables other tools to do proper service discovery and provenance logging.

@@

We list a selection of the most important software included in LaMachine, grouped by research institute:

\begin{itemize}
 \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University
     Nijmegen\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
    \url{https://languagemachines.github.io/ucto}, \url{https://languagemachines.github.io/frog},
    \url{http://ilk.uvt.nl/wopr}, \url{https://proycon.github.io/folia}, \url{https://proycon.github.io/clam},
\url{https://github.com/proycon/flat}, \url{https://proycon.anaproy.nl/pynlpl},
\url{https://proycon.github.io/colibri-core/},
\url{https://github.com/proycon/gecco},
\url{https://github.com/proycon/valkuil-gecco},
\url{https://github.com/LanguageMachines/PICCL}, \url{https://github.com/LanguageMachines/ticcltools},
\url{https://github.com/proycon/labirinto},
\url{https://github.com/proycon/oersetter-webservice}
}
 \begin{itemize}
     \item \textbf{Timbl} - A memory-based machine learning
         toolkit, and \textbf{Mbt}, a memory-based tagger based on timbl.
     \item \textbf{Ucto} - A tokeniser
     \item \textbf{Frog} - Frog is an integration of various memory-based natural language processing (NLP) modules
         developed for Dutch. It can do Part-of-Speech tagging, lemmatisation, named entity recogniton, shallow parsing,
         dependency parsing and morphological analysis. Also included in LaMachine;
         \textbf{Toad}, Trainer Of All Data, training tools for Frog.
     \item \textbf{Wopr} - Memory-based Word Predictor.
     \item \textbf{CLAM} - Quickly build RESTful webservices, powers many webservices offered by LaMachine.
     \item \textbf{FoLiA} - Format for Linguistic Annotation, with tools and libraries in/for Python and $C++$.
     \item \textbf{FLAT} - FoLiA Linguistic Annotation Tool: a web-based linguistic annotation tool.
     \item \textbf{PyNLPl} - Python Natural Language Processing Library.
     \item \textbf{Colibri Core} -- Colibri core is an NLP tool as well as a C++ and Python library for working with
         basic linguistic constructions such as n-grams and skipgrams (i.e patterns with one or more gaps, either of
         fixed or dynamic size) in a quick and memory-efficient way. -
     \item \textbf{Gecco} - Generic Environment for Context-Aware Correction of Orthography - and \textbf{Valkuil.net},
         an instantiation thereof for Dutch.
     \item \textbf{PICCL} - PICCL: A set of workflows for corpus building through OCR, post-correction (through \textbf{TICCL}) and Natural Language Processing.
     \item \textbf{Labirinto}- A web-based portal listing all
         available tools in LaMachine, an ideal starting point for LaMachine.
     \item \textbf{Oersetter} - A Frisian-Dutch Machine Translation system in collaboration with the Fryske Akademy.
 \end{itemize}
 \item by the University of Groningen
 \begin{itemize}
     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - a dependency parser and tagger for Dutch
 \end{itemize}
 \item by the Vrije Universiteit Amsterdam
 \begin{itemize}
     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files
 \end{itemize}
 \item by Utrecht University
 \begin{itemize}
     \item  \textbf{T-scan}\footnote{\url{https://github.com/proycon/tscan}} - T-scan is a Dutch text analytics tool for readability prediction (initially developed at TiCC, Tilburg University, and in collaboration with Radboud University, Nijmegen)
 \end{itemize}
 \item by Meertens Instituut
 \begin{itemize}
     \item \textbf{Python Course for the Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} - Interactive
         tutorial and introduction into programming with Python for the humanities
 \end{itemize}
\end{itemize}

In addition to the above listed specific software, LaMachine also incorporates a large number of renowned tools by external international parties, offering
most notably a mature Python environment with renowned scientific modules. The following list is not
exhaustive: \footnote{Links: \url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
    \url{https://spacy.io},  \url{http://www.nextflow.io}, \url{https://stanfordnlp.github.io/CoreNLP/},  \url{https://github.com/tesseract-ocr/tesseract}, \url{https://tensorflow.org},
\url{http://kaldi-asr.org}, \url{http://www.statmt.org/moses}}

\begin{itemize}
    \item \textbf{Python}: Numpy, Scipy, Matplotlib, Scikit-Learn, IPython, Jupyter, ...
    \begin{itemize}
        \item \textbf{Jupyter Lab} The successor of the popular Jupyter Notebooks, offers notebooks, a web-based IDE, terminals. An ideal entry point to get started with LaMachine and all it contains!
        \item \textbf{PyTorch} - Deep-learning library for Python
        \item \textbf{NLTK} - Natural Language Toolkit for Python
        \item \textbf{Spacy} - Industrial-Strength NLP in Python
    \end{itemize}
    \item \textbf{R}
    \item \textbf{Java}
    \begin{itemize}
        \item \textbf{NextFlow} - A system and language for writing parallel and scalable pipelines in a portable manner.
        \item \textbf{Stanford CoreNLP} - Various types of linguistic enrichment
    \end{itemize}
    \item \textbf{Tesseract} - Open Source Optical Character Recognition (OCR)
    \item \textbf{Tensorflow} - Open-source machine learning framework
    \item \textbf{Kaldi} - Speech Recognition Framework (ASR)
    \item \textbf{Moses} - Statistical Machine Translation system
\end{itemize}

@@

In any of its flavours, LaMachine offers low-level shell access, i.e. a command-line interface accessed through a
terminal. In flavours that are separated from the host system by a network, this is accomplished over \texttt{ssh}.
After accessing the LaMachine environment through a terminal, as shown in Figure~\ref{fig:venv}, the user has the liberty to do whatever she wants and can invoke any
of the tools that offer a text interface, including text editors such as vim, emacs or nano, version control systems
such as git and interpreters such as Python or R (or compilers for that matter). The latter in turn allow users to use
one of the many specialised programming libraries included in LaMachine to build their own tools. All this makes
LaMachine ideally suited as a development environment.

@@

LaMachine comes with a webserver\footnote{In situations where web interfaces are not needed or desired, the
user has the ability to opt-out of this.}. This webserver serves various web-capable tools that are incorporated in LaMachine. One
of these tools is a portal website that provides an overview of all installed software, and acts as a point of access to
all its web services and web applications. This portal, as shown in
Figure~\ref{fig:portal}, leverages the metadata registry compiled in each LaMachine installation (see
section~\ref{sec:metadata}).

@@

LaMachine also comes with a
Jupyter Lab\footnote{\url{https://jupyter.org/}} installation which provides a web-based Integrated Development
Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
text, code and data output and have gained great popularity in the data science community. This is shown in
Figure~\ref{fig:lab}. This type of virtual laboratory provides a powerful interface that provides an alternative to
regular shell access and may have a larger appeal for those parts of the audience that do not feel completely
comfortable in the terminal.

@@

LaMachine provides various CLAM-based webservices. CLAM is a long-time CLARIN project that allows developers to turn
their command-line NLP tool into a RESTful webservice with relatively little effort \cite{CLAMPAPER}, and which in
addition also provides a generic web-interface for human end-users. The portal website in LaMachine links to these. This
again addresses an audience that does not require a high degree of technical expertise, but is suitable for a wider
audience of researchers, including those in the Humanities.

@@

Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. But that
would not be appropriate as we characterise LaMachine more as a meta-distribution which simply builds on the foundation
of several existing major Linux distributions. As a meta-distribution that runs in a variety of flavours and on a
variety of platforms, it offers a great deal of flexibility that normal distributions do not have. At the same time, the scope of
LaMachine is more narrowly defined than that of a generic Linux distribution, including even specialised Linux
distributions that focus on scientific computing, such as Fedora
Scientific\footnote{\url{https://labs.fedoraproject.org/en/scientific/}}.

@@

At the start of the development cycle for LaMachine v2, we investigated whether Anaconda and its ecosystem would provide
a sufficient foundation for us to build upon. We concluded this was not the case for a number of reasons: Anaconda
introduces more overhead than we desired for our purposes and conflicted with certain technology choices we made. We
wanted our native flavours (local environment and global environment) to be as close to the underlying distribution as possible and to reuse existing
technologies and packages, such as compilers (e.g. gcc) and
interpreters (e.g. python), from the distribution (or from homebrew for macOS).  Anaconda, on the other hand, choses to provide its own packages and builds on those. It
does so using its own package manager (conda) and package repositories (conda-forge). This would require repackaging certain software for the
anaconda ecosystem.
